%==========================================================
\section{Abstract}\label{sec:ex1}
%==========================================================
In this project, I have studied three regression methods, Ordinary Least Squares (OLS), Ridge and Lasso, and compared their performance on synthetic and real terrain data. I found that the choice of method relies on the problem to be solved, and that the OLS method is sufficient when using noise free synthetic data from the Franke function. When introducing noise to the function, Ridge regression performed better as it did not overfit the data. I also performed a bias-variance trade-off analysis, using resampling methods such as bootstrap and cross-validation, and found the optimal polynomial degree for the input data.

%==========================================================
\section{Introduction}\label{sec:ex2}
%==========================================================
Artificial Intelligence (AI) and Machine Learning (ML) methods have many use cases, as it allows us to automate manual processes, make sense of complex data, and learn patterns not readily available. Today, the use of ML methods can be found in most industries, such as fraud detection, diagnosing disease, and increase efficiency in manufacturing processes \cite{forbes:2023:machine_learning}. 

In this project I will focus on Linear Regression ML methods, and their use in modeling and analysing terrain data. More specifically, I will study the difference in performance of Ordinary Least Squares, Ridge, and Lasso regression method, on synthetic terrain data generated using the Franke function. In addition, I will compare the model's performance when testing on real terrain data. To analyse the models in more detail, I will use resampling methods such as bootstrap and cross-validation, and study the bias-variance trade-off.

In Section II, I will present the methods and tools used in this project. Continuing with Section III, where I show the results and discuss my findings. In Section IV, I conclude and discuss possible future work.


% It is a growing field of vast potential. such as insurance and banking, 
% , which can be important in detecting changes in natural terrain